---
title: 'CHL5202: Biostatistics II'
output: 
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
---


# Course Description and Objectives {.unlisted .unnumbered}
* The goal of this course is to continue development and application of skills in statistical analysis required in epidemiologic and health care research.
* Analysis where outcome is continuous and exposure and covariates are a mix of continuous and categorical variables (multiple linear regression)
* Relaxing linearity assumptions of continuous exposure variables through the use of regression splines
* Perform internal validation by means of bootstrapping or cross-validation
* Extension to binary outcomes (multiple logistic regression)
* Extension to time to event and censored outcomes (survival analysis)
* Dealing with missing data
* Introduce Bayesian methods

Course material provided by Professor Kevin Thorpe and modifications were made by myself.

# Tutorial 1

## Question 1
Load the data frame `tut1` into R from the file `tut1.RData` available on the course page. The data
frame contains the variables x and y.

Fit a simple regression with y as the response and x as the predictor. Use diagnostic plots to determine if a non-linear relationship should be modeled. If you believe a non-linear relationship is present, model it using a restricted cubic spline. The number of knots is up to you.

```{r, message=F}
# Load packages and data
library(rms)

load("./CHL5202/tut1.RData")

dd <- datadist(tut1)
options(datadist="dd")
```

* Some notes taken from [here](https://thomaselove.github.io/432-notes/using-ols-from-the-rms-package-to-fit-linear-models.html).

* Comparing `ols()` vs `lm()`: one advantage for ols is that the entire variance-covariance matrix is saved

```{r}
fit0 <- ols(y~x,data=tut1)
fit0
```

* When applying anova to an ols model, it separates out the linear and non-linear components of restricted cubic splines and polynomial terms (and product terms if they are present). 
* Unlike for `lm()`, using `anova` with `ols()` we have partial F-tests that describe the marginal impact of removing each covariate from the model.

```{r}
anova(fit0)
```

* Next, we check the model assumptions using diagnostic plots. 
* In the Normal Q-Q Plot we are looking for normally distributed error terms, i.e. most of the points lie on the line and do not show a non-linear pattern. 

```{r}
# for diagnostic plots
dx0 <- data.frame(tut1, e = resid(fit0), yhat = fitted(fit0)) 

# Normal QQ Plot
qqnorm(dx0$e)
qqline(dx0$e)
```

* `xyplot()`: Common Bivariate Trellis Plots. Produces conditional scatterplots, typically between two continuous variables. Below we have residual plots. 
* We are looking for the residuals to not show a distinct (non-linear) pattern which would suggest that a transformation is required.

```{r}
# Residual Plots
xyplot(e~x, data = dx0, type = c("p","smooth"))
xyplot(e~yhat, data = dx0, type = c("p","smooth"))
```

* A non-linear relationship is present, so we will model it using a restricted cubic spline. For this example, we are going to use 4 knots.

```{r}
fit1 <- ols(y ~ rcs(x, 4), data = tut1)

fit1
```

* It is difficult to interpret the coefficients above. The `anova()` output below is more useful and shows a significant non-linear component.

```{r}
anova(fit1)
```


```{r}
# Plot predicted fit with confidence interval
ggplot(Predict(fit1))
```


```{r}
# for diagnostic plots
dx1 <- data.frame(tut1, e = resid(fit1), yhat = fitted(fit1))

# Normal QQ Plot
qqnorm(dx1$e)
qqline(dx1$e)
```


```{r}
# Residual plots
xyplot(e ~ x, data = dx1, type = c("p", "smooth"))
xyplot(e ~ yhat, data = dx1, type = c("p", "smooth"))
```

* There is no longer a non-linear pattern in the residuals.


## Question 2

Load the data frame FEV from the file FEV.RData. For these data, use the variable fev as the response and the rest as the explanatory covariates.

Fit an additive linear model to fev using the other variables as the covariates. Evaluate whether any of the continuous variables should be fit as non-linear terms. 

Fit an another additive model but this time model the continuous covariates with restricted cubic splines with four knots. Describe the results.

```{r}
load("./CHL5202/FEV.RData")

fev.dd <- datadist(FEV)
options(datadist = "fev.dd")
```


```{r, message=F}
# `dplyr()`: package for data wrangling - part of `tidyverse()`
library(dplyr)

# Useful for quickly viewing the data
glimpse(FEV)

summary(FEV)
```


```{r}
# Fit initial additive model
fev.lin <- ols(fev ~ age + height + sex + smoke, data = FEV)
fev.lin
```

```{r}
anova(fev.lin)
```

* `age` and `height` are continuous variables, so we can use residual plots.
* `sex` and `smoke` are categorical variables (in fact, they're binary) so residual plots won't be useful.

```{r}
dx.fev <- data.frame(FEV, e = resid(fev.lin), yhat = fitted(fev.lin))

xyplot(e ~ age, data = dx.fev, type = c("p", "smooth"))
xyplot(e ~ height, data = dx.fev, type = c("p", "smooth"))
```

* The residual plots show that non-linearity needs to be modelled for both variables.
* Once again, we shall use restricted cubic splines with 4 knots.

```{r}
fev.full <- ols(fev ~ rcs(age, 4) + rcs(height, 4) + sex + smoke, data = FEV)

fev.full
```

```{r}
anova(fev.full)
```

* The `anova()` output suggests that the non-linear components for age and height are statistically significant.

```{r}
dx.full <- data.frame(FEV, e = resid(fev.full), yhat = fitted(fev.full))

# Residual plots look good
xyplot(e ~ age, data = dx.full, type = c("p", "smooth"))
xyplot(e ~ height, data = dx.full, type = c("p", "smooth"))
```

* Running `summary()` is the best way to get the effects of the categorical variables.
* For continuous variables, it is better to plot.

```{r}
summary(fev.full)
```

```{r}
ggplot(Predict(fev.full, age))
ggplot(Predict(fev.full, height))
```


## Extra

* `%>%`: pipe operator - very useful for performing multiple data wrangling steps
* `ggplot2()`: plotting package, included in `rms()` and `tidyverse()`

```{r}
# Boxplot
FEV %>%
  ggplot(mapping = aes(x = sex, y = height, fill = sex)) +
  geom_boxplot() +
  theme_bw() +
  labs(title = "Boxplot of heights by sex")
```

