---
title: "STA238: Probability, Statistics and Data Analysis II"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
---


# Course Description {.unlisted .unnumbered}

An introduction to statistical inference and practice. Statistical models and parameters, estimators of parameters and their statistical properties, methods of estimation, confidence intervals, hypothesis testing, likelihood function, the linear model. Use of statistical computation for data analysis and simulation.

Course material provided by Professor Karen Huynh Wong and modifications were made by myself.

# Lab 0

In this R Lab, you will be guided by your TA to familiarize yourself with the R environment, using and formatting in R Markdown, and producing well-organized outputs. 


## R Lab Learning Goals

  * Install and load packages in R, use them in an R Markdown file
  * Simple computations and referring to `?command_name` to find the documentation of various commands
  * Select elements, columns, or rows from a dataset
  * Produce simple plots in ggplot with informative and clearly formatted titles + labels
  * Use LaTeX notation to neatly and correctly display math notation. You can knit this document as you type (which also helps you identify bugs in your code!) to see how the format of the text translates into the finished product, such as creating headers, bold-facing, and math notation!
  * Knit your document to a finished document that includes written text and properly rendered math notation, with plots and source code separately embedded

Some additional resources you may find helpful as you learn is the [R Markdown Cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf) and the [LaTeX Cheatsheet](http://tug.ctan.org/info/undergradmath/undergradmath.pdf). As you work more in LaTeX, the notation will get easier and become second nature!


## 1. Install and Load Packages

R packages are a collection of functions, code, and data sets that are developed by others and can be used to supplement the basic tools in R. R comes with some basic packages, but working with them can be sometimes clunky. You'll begin by learning how to install packages and load packages, and the difference between these two actions!

Begin by installing and loading the `tidyverse` package, which is a collection of R packages. We will use this extensively in our course, as all packages within share similar structure and design, making it a very intuitive and versatile tool to master!

**Note: Never include a package install in your R chunks in an R Markdown file. R will not be able to knit your document! Instead, run it strictly within your console.**

```{r, message=F}
# Students: What do you think the 'message = F' in the R chunk option does? 
# Try knitting your document with and without the option

# Can also try `warning = F` to hide warnings
# echo = F` to run the chunk but hide the code

# install.packages("tidyverse")
library(tidyverse)

# Use Ctrl + Alt + I shortcut for creating code chunks
```

## 2. Loading Data and Saving it to an Object

Let's examine some datasets that are already available in R (see `library(help = "datasets")` for a list of datasets in base R), starting with the ToothGrowth dataset. 

```{r}
# ?ToothGrowth

# Variable Assignment
dat <- ToothGrowth

# Mostly useful for smaller datasets, but for larger ones we prefer to use glimpse and summary
# View(dat)

glimpse(dat)
summary(dat)
```

## 3. Simple Data Filtration

It looks like there are two groups of data in the `ToothGrowth` data set. What are they? 

What if you only want to study the growth based on one source of vitamin C? How do you go about extracting this information?

```{r, eval=F}
# After loading and examining your data, and learning about selecting columns or elements
# from your data set, consider the following two commands below. 
# Briefly comment below what these two commands do:

# What does
which(dat$supp == 'VC') #do?
# Extracts which indices are TRUE
```


```{r, eval=F}
# What does 
dat[dat$supp=='VC', 1] #do?
# extracts the `len` variable/column, where `supp` is 'VC'

# Alternative 1
# Pipe operator shortcut: Ctrl + Shift + M
dat %>% 
  filter(supp == "VC") %>% 
  select(len)

# Alternative 2
# This is basically Alternative 1 without the pipe operator
select(filter(dat, supp == "VC"), len)
```


```{r, eval=F}
# Below we can also extract the other 2 columns
dat[dat$supp=='VC', 2]
dat[dat$supp=='VC', 3]

```

## 4. Formulating a Question

After examining the data set, what are some study questions we can ask? Make some suggestions in the Teams chat!

Below are some questions suggested by students in the chat:
* Does dosage and length have a positive correlation?
* Which vitamin contributes to the most growth?

## 5. Calculations in R

Find the (sample) mean and standard deviation of tooth length in the Vitamin C group and again in the group that only received Orange Juice. What can be said based on the numbers computed about the differences between the two groups? 

Do this using the formula, instead of the functions `mean()` or `sd()` to practice using R as a calculator!

```{r}
# Task for students: try to recreate the mean and variance values through
# Calculations in R, and compare them with the values you should get through
# mean() and sd() commands in R

# Create tooth length variable for the vitamin C 
vc <- dat[dat$supp=='VC', 1]

# take the sum of the tooth length, then divide by the number of values
n <- length(vc)
mean.vc <- sum(vc)/n
mean.vc

# Can also compute using a for loop
mean.vc2 <- 0
for (i in 1:n){
  mean.vc2 <- mean.vc2 + vc[i]/n
}
mean.vc2

# Compare to mean() function
mean(vc)

# Standard deviation
var.vc <- sum((vc - mean.vc)^2)/(n-1)
sd.vc <- sqrt(var.vc)
sd.vc

# Compare to sd() function
sd(vc)
```

Below is a repeat of the above but for `OJ` (orange juice):

```{r}
# Orange Juice
oj <- dat[dat$supp=='OJ', 1]

# Mean
n <- length(oj)
mean.oj <- sum(oj)/n
mean.oj

# Can also compute using a for loop
mean.oj2 <- 0
for (i in 1:n){
  mean.oj2 <- mean.oj2 + oj[i]/n
}
mean.oj2

# Compare to mean()
mean(oj)


# Standard deviation
var.oj <- sum((oj - mean.oj)^2)/(n-1)
sd.oj <- sqrt(var.oj)
sd.oj

# Compare to sd()
sd(oj)
```

Trying to type the variance formula in plain-text is not visually appealing, nor helpful in communicating (it's very difficult to read!). The beauty of LaTeX is the format is fairly intuitive, such as `$\frac{numerator}{denominator}$` will render as $\frac{numerator}{denominator}$ (the dollar signs indicate the start and end of the math setting, double-dollar sign will have the math rendered centre to the page). For example, the sample variance can be written as:

$$s^2 = \frac{\sum_{i=1}^n (x_i - \bar{x})^2}{n-1}$$
Try to write the sample mean in LaTeX and quote the values computed in the R chunk above!

For the mean we can use `\bar{x}_n = \frac{\sum_{i=1}^n x_i}{n}`:
$$\bar{x}_n = \frac{\sum_{i=1}^n x_i}{n}$$

## 6. Simple Plots 

Numerical summaries can be powerful to provide concrete, numerical evidence to support our claims. Using graphical displays can also communicate clearly in a visual manner the story that our numbers tell, and can sometimes be a more powerful communication tool than numbers because we can *see* the *scale* of differences. Let's use the `ggplot` package in `tidyverse` to produce side-by-side box plots of how tooth length growth differs between the two sources of vitamin C: 

```{r, fig.align='center', fig.dim=c(7, 5)}
# Focus of this segment: understanding the layering language of ggplot
# Note also the R chunk options that control the alignment and size of
# The output figures

dat %>% 
  ggplot(aes(x = supp, y = len)) +
  geom_boxplot() + # passing aes(x,y) into here works too!
  labs(title = "Guinea Pig Tooth Length Growth by Vitamin C Source",
       subtitle = "ToothGrowth Dataset",
       x = "Vitamin C Source",
       y = "Tooth Length Growth") +
  scale_x_discrete(labels = c("Orange Juice", "Ascorbic Acid")) +
  theme_bw() # there are many different themes from which to choose!
```

What can we notice about tooth length growth between the two supplement groups? 

Practice using visual and numeric features to:

  1. Write a conclusion about any perceivable differences between the two sources of vitamin C.
  2. Provide graphical and numerical features and their <span style="text-decoration:underline">interpretations</span>, explaining how they support your claim(s)!

Here are things that we notice:

* The median tooth length growth for the orange juice supplement is almost at same level as the third quartile of tooth length growth for ascorbic acid.  
* There is some evidence which suggests that median tooth length growth for the orange juice supplement is statistically significantly larger than for the ascorbic acid supplement.
* We can use a t-test to test for the difference in means between two groups. This is just a quick example but make sure to check the assumptions before proceeding!

```{r}
# Two-sided t-test
t.test(oj,vc)

# One-sided t-test
t.test(oj,vc,"greater")
```

* From the two-sided t-test (p-value = 0.06), there is some evidence to suggest that the difference in tooth length growth between the two supplement groups is statistically significant. 
* From the one-sided t-test (p-value = 0.03), there is evidence to suggest that the mean tooth length growth for guinea pigs with orange juice supplements is statistically significantly greater than the mean tooth length growth for guinea pigs with ascorbic acid supplements.
* Note that the p-value for the one-sided t-test is half the p-value for the two-sided t-test.


# Lab 1
In this R Lab, you will be guided by the TA to demonstrate the steps involved in performing an exploratory data analysis.


## R Lab Learning Goals

  * Load data from a local csv file
  * Formulate a question for EDA
  * Examine and pre-process your data using functions from the `tidyverse` package
  * Use the `lubridate` package for manipulating dates
  * Produce numerical summaries using the `summarize` function
  * Produce plots in ggplot to explore your EDA question


```{r, message=F, warning=F}
# Load Required Packages
library(tidyverse)
library(lubridate)

# Set work directory
setwd("~/GitHub/dang-kevin.github.io/STA238")

# Read in Data
weather <- read.csv("weatherstats_vancouver_daily.csv")

# Using file.choose() allows you to manually select the file you want
# weather <- read.csv(file.choose())

# Quick overview of the data
glimpse(weather)
```

Note that since this data is from a csv file, functions like `?weather` will not provide us with information about the data set. For the purposes of this lab, we assume that the variable names are sufficiently informative to intuit their contents. In practice, however, information about the data and its variables can be found in a documentation file that is usually available from the source of the data.

## 1. Formulating questions for EDA

Our goal during exploratory data analysis (EDA) is to develop an understanding of our data. Real data is complicated, often containing missing data and irregular formatting.

In this lab, we will demonstrate this process with the question:
**"What is the monthly total rainfall in Vancouver over 2021?"**

## 2. Checking the `rain` variable

The variable `rain` contains the daily rainfall measured in millimeters. Our first task is to check for any missing data. We could do this using the base R commands from Lab 0, but the code gets messy for complicated manipulations. Instead, we will use the pipe operator `%>%` from the `tidyverse` package along with some data processing functions. The pipe operator `%>%` allows us to take the output from one function and "pipe it in" to the argument of the next function. You will find that data manipulations using the pipe operator are much more readable than the equivalents in base R!

```{r}
# Examples of pipe operator

#1. Pulling the first 5 rows of data containing date, rain, and average dew point

# Tidyverse way
weather %>% 
  select(date, rain, avg_dew_point) %>%
  head(5)

# base R way
head(weather[c("date", "rain","avg_dew_point")], 5)
# head(weather[, c("date", "rain","avg_dew_point")], 5)

#2. Adding new columns of data to the existing weather data:

# Tidyverse way
weather <- weather %>% 
  mutate(new_variable_1 = 1.5 * avg_hourly_temperature,
         new_variable_2 = rain + 10)

# base R way
weather$new_variable_1 <- 1.5 * weather$avg_hourly_temperature
weather$new_variable_2 <- weather$rain + 10
```

First we will check which dates have missing rainfall data:
```{r}
# Base R way
weather$date[is.na(weather$rain)]

# Tidyverse way
weather %>% 
  filter(is.na(rain)) %>% 
  # can also use select(date)
  pull(date)

# To show the days that DO NOT have missing rain values, add '!' in front of is.na()
# e.g. weather$date[!is.na(weather$rain)]
```

We see that there is one missing value of rainfall in the year 2021. This might be a cause for concern if this was a day with particularly high rainfall. Unfortunately, we do not know if this was the case. The best we can do in these situations is to tell the reader that what is missing from our data so that they are aware of the limitations of our analysis.

## 3. Formatting the `date` variable

Since we are concerned with total monthly rainfall, our next task is to convert the `date` variable into a form that is amenable to analysis. First we examine the format `date`:

```{r}
# str() compactly displays the structure of an object
# Good alternative to glimpse()
str(weather$date)
```

We find that each entry of `date` is a character vector of the form "YYYY-MM-DD". To prepare this data for use in `lubridate`, we convert `date` from a character vector to a datetime object. 

```{r}
# First we convert date to a datetime object
# To explain format, look at `?strptime`

# The base R way
weather$date <- as.Date(weather$date, format = "%Y-%m-%d")

# The tidyverse way using mutate()
# Make sure to reassign the dataset to ensure that the mutation step is saved
weather <- weather %>% 
  mutate(date = as.Date(date, format = "%Y-%m-%d"))
```

Now that we have a datetime object, we can use functions from `lubridate` to easily convert `date` into many useful quantities. We use the functions `year`, `month`, and `day` to define the corresponding variables in our data

```{r}
######################################################################
###  NOTE: There is an 'eval=F' in the R Chunk Option! Make sure to ##
###        delete this before knitting!!                            ##
######################################################################

# Demonstrate some lubridate functions

# Extract weekday as character
weekdays(weather$date) %>% str()
# Extract month as integer
lubridate::month(weather$date) %>% str()
# Extract month as ordered factor
month(weather$date, label = TRUE) %>% str()
# Extract the year as integer
year(weather$date) %>% str()

# summarizing the pieces above
# Create the year, month, and day variables
weather <- weather %>%
  mutate(year = year(date),
         month = month(date, label = TRUE),
         day = day(date))
```

## 4. Filtering and summarizing the data

Now that we have examined and processed our data, we can start to answer our original question: "What is the monthly total rainfall in Vancouver over 2021?" To begin, we might use `filter` and `select` to inspect our data

```{r}
weather %>% 
  filter(year == 2021) %>% 
  select(year, month, day, rain) %>% 
  head(10)
```

This output, however, is not as helpful as a numerical summary of the data. The function `group_by` and `summarize` provide a useful method for summarizing the data. First we choose a variable(s) to group by and then we define the numerical summaries we wish to compute. For example, we might be interested in the mean and the maximum rainfall each month.

```{r}
# Assign to new dataframe to use for plotting
df <- weather %>% 
  filter(year == 2021) %>% 
  group_by(month) %>% 
  summarise(rain_tot = sum(rain, na.rm = T),
            rain_max = max(rain, na.rm = T))

df
```

These numerical summaries may be useful in some contexts, but a graph is generally better for examining these summaries. Thankfully, the results of the summary can easily be piped into ggplot to make such graphs!

```{r, fig.width=8, fig.height=5}
df %>% 
  ggplot(aes(x = month, y = rain_tot)) +
  geom_col(fill = "skyblue") +
  labs(title = "Total monthly rainfall, Vancouver 2021",
       x = "Month",
       y = "Total rainfall (mm)") +
  # Alternative labelling method instead of labs()
  # ggtitle("Total monthly rainfall, Vancouver 2021") +
  # xlab("Month") +
  # ylab("Total rainfall (mm)") +
  theme_bw()
```

Some quick observations:

  * No rainfall in July.
  * A lot of rain in November! 
